EGGs are adhesive!

### 2025-03-12 calco/main.tex
Sezione 4.2.

\noindent
Thus, $L$ and $R$ are the lift- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet they are not affected by the rule itself.
%
If a category is adhesive, and the arrows of the rules are monomorphisms, then the presence of a match ensures that the
two pushouts exist, hence a rewriting step can be performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{xxx} extend the core framework, still ensuring that if the arrow 
of the rules are in $\mathcal{M}$,
%and the match is in $\mathcal{N}$, 
then the classical results from the theory of graph 
transformation can still be lifted~\cite{xxx}. 
If only the left-hand side belongs to $\mathcal{M}$, the current theory is stil under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness and the immediate advantages of proving a category 
to be adhesive, proving that a given category satisfies the conditions 
for being adhesive can be a daunting task. For this reason,  sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some graph-like structures, such as those called \emph{hypergraph with equivalence} in~\cite{concur2006}, the question 
of their adhesivity has not yet been settled.

@@

\begin{restatable}{lemma}{kpp}\label{lemma:kern_pairs_pres_pullbacks}
	Let $f\colon X \to Y$ and $g\colon Z \to W$ be two arrows admitting kernel pairs and suppose that the solid part of the three squares below is given. 
	%
	If the leftmost square is commutative, then there is a a unique arrow $k_h\colon K_f \to K_g$ making the other three commutative.
	\[\xymatrix@R=15pt{X \ar[r]^{h}\ar[d]_{f}& Z \ar[d]^{g} & K_f \ar@{.>}[r]^{k_h} \ar[d]_{\pi^1_f}& K_g \ar[d]^{\pi^1_g} & K_f  \ar@{.>}[r]^{k_h} \ar[d]_{\pi^2_f}& K_g \ar[d]^{\pi^2_g}& K_f  \ar@{.>}[r]^{k_h} \ar[d]_{(\pi^1_f,\pi^2_f)} & K_g \ar[d]^{(\pi^1_g, \pi^2_g)}\\ Y \ar[r]_t& W & X \ar[r]_{h} & Z & X\ar[r]_{h} &Z & X\times X \ar[r]_{h\times h}& X\times Z}\]

		Moreover, the following hold true:
		\begin{enumerate}
			\item if $h$ is mono then $k_h$ is mono too;
			\item if the leftmost square is a pullback, then also the central too are so;
			\item if $h$ is mono and the first square is a pullback then the last one is a pullback too.
		\end{enumerate} 
\end{restatable}
%[Proof in \Cref{kpp-proof}]

@@

\noindent
\parbox{7.5cm}{\begin{restatable}{lemma}{mps}\label{prop:kerset}
	Suppose that in $\Set$ the cube in the diagram on the left is given, whose top and bottom faces are pushout, the left face a pulllback and such that $m\colon A\mto C$ is an injection. Then the following are true:
	\parbox{14cm}{\begin{enumerate}
		\item the right face of the cube is a pullback too;
		\item the right square, made by the kernel pairs of the vertical arrows, is a pushout.
	\end{enumerate}}
\end{restatable}}\hfill 
\parbox{6cm}{\vspace{-1.1cm}\xymatrix@C=10pt@R=5pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar@{>->}[dl]^{m} && B \ar@{>->}[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar@{>->}[rr]_{g} & & D }}

@@

\begin{corollary}\label[corollary]{prop:hypadh}
	 $\hyp$ is an adhesive category.
\end{corollary}
%\begin{proof}
%	By hypothesis $(-)^\star$ preserves pullbacks, while $\pro$ is continuous by definition, thus the thesis follows from  \Cref{thm:slice-functors} and \Cref{prop:com}.   
%\end{proof}

@@

\begin{definition}
	A \emph{hypergraph with equivalence} $\mathcal{G} = (E_\mathcal{G}, V_{\mathcal{G}}, Q_\mathcal{G}, s_\mathcal{G}, t_\mathcal{G}, q_\mathcal{G})$ is a 6-tuple such that $\mathcal{G} = (E_\mathcal{G}, V_{\mathcal{G}}, s_\mathcal{G}, t_\mathcal{G})$ is a hypergraph, $Q_\mathcal{G}$ is a set and $q_{\mathcal{G}}: V_{\mathcal{G}}\eto Q_{\mathcal{G}}$ is a surjection called \emph{quotient map}. 
	%
	A morphism $h\colon \mathcal{G\to H}$ is a triple $(h_E, h_V, h_C)$ such that the following diagrams commute
	\[\xymatrix@R=15pt{
		{E_\mathcal{G}}\ar[r]^{s_\mathcal{G}}\ar[d]_{h_E} & {V_{\mathcal{G}}^\star}\ar[d]^{h_V^\star} & {E_\mathcal{G}}\ar[r]^{t_\mathcal{G}}\ar[d]_{h_E} & {V_{\mathcal{G}^\star}}\ar[d]^{h_V^\star} & {V_\mathcal{G}}\ar@{>>}[r]^{q_\mathcal{G}}\ar[d]_{h_V} & {Q_{\mathcal{G}}} \ar[d]^{h_C} \\
		{E_\mathcal{H}}\ar[r]_{s_\mathcal{H}} & {V_{\mathcal{H}}^\star}	& {E_\mathcal{H}}\ar[r]_{t_\mathcal{H}} & {V_{\mathcal{H}^\star}}& {V_\mathcal{H}}\ar@{>>}[r]_{q_\mathcal{H}} & {Q_\mathcal{H}}
	}\]
	The category of hypergraphs with equivalences and their morphisms is denoted $\EqHyp$.

@@

\begin{definition}
	Let $\Sigma$ be an algebraic signature and $\mathcal{G}^{\Sigma}$ the hypergraph associated to it. A \emph{labelled hypergraph with equivalence} is a pair $(\mathcal{H}, l)$ where $\mathcal{H}$ is an object of $\EqHyp$ and $l$ a morphism $T(\mathcal{H})\to \mathcal{G}^\Sigma$ of $\EqHyp$. A \emph{morphism of labelled hypergraphs with equivalence} between $(\mathcal{H}, l)$ and $(\mathcal{H}', l')$ is an arrow $h\colon \mathcal{H}\to \mathcal{H}'$ such that $l= l'\circ T(\mathcal{H})$.

@@

In general, note that both a maximally minimally and a shared representation of an EGG exist, and for our 

@@

What about rewriting? The theory of $\mathcal{M}$-adhesivity ensures that if the rules are spans of 
arrows in $\mathcal{M}$, then we can lift all the standard properties
of the DPO approach.
% to the category at hand. 
However, the structure of rules that is advanced in~\cite{DetlefsNS05},
is given as a span $L \leftarrow L \rightarrow R$, where the first component is the identity, hence in $\mathcal{M}$,
while the second component may not belong to $\mathcal{M}$.

\noindent
\begin{minipage}[l]{.48\linewidth}
Consider e.g. an EGG rule such $x / x \to 1$, from the introductory example in~\cite{WillseyNWFTP21}.
It can be modelled as the DPO rule aside, concisely represented as the arrow $L \rightarrow R$,
which can be applied with a match that is injective on equivalence classes to the legal EGGs
above.
\end{minipage}\hfill
\begin{minipage}[r]{.60\linewidth}
\xymatrix{        
 \begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
                \node[style=node](v1) at (-0.9, 0.3){};
                \node[style=node](v2) at (-0.9, -0.3){};
                \node[style=medium box] at (0, 0){$/$};
                \node[style=none](divfst) at (-0.3, 0.3){};
                \node[style=none](divsnd) at (-0.3, -0.3){};
                \node[style=none](divout) at (0.3, 0){};
                \node[style=node] (w) at (0.9, 0){};
        \end{pgfonlayer}        
        \begin{pgfonlayer}{edgelayer}
                \draw(v1) to (divfst.center);
                \draw(v2) to (divsnd.center);
                \draw(divout.center) to (w);
        \end{pgfonlayer}
        \begin{pgfonlayer}{eqlayer}
                \draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
                \end{pgfonlayer}\begin{pgfonlayer}{background}
                        \draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
                \end{pgfonlayer}
        \end{tikzpicture}
	\ar@{=>}[r] &
        \begin{tikzpicture}[baseline=(nil.center)]\begin{pgfonlayer}{nodelayer}
                \node[style=node](v1) at (-0.9, 0.3){};
                \node[style=node](v2) at (-0.9, -0.3){};
                \node[style=medium box] at (0, 0){$/$};
                \node[style=none](divfst) at (-0.3, 0.3){};
                \node[style=none](divsnd) at (-0.3, -0.3){};
                \node[style=none](divout) at (0.3, 0){};
                \node[style=node] (w) at (0.9, 0){};
                \node[style=small box] at (0, -1){$1$};
                \node[style=none](one) at (0.3, -1){};
                \node[style=node](z) at (0.9, -1){};
                \node[style=none](nil) at (0, -0.5){};
        \end{pgfonlayer}        
        \begin{pgfonlayer}{edgelayer}
                \draw(v1) to (divfst.center);
                \draw(v2) to (divsnd.center);
                \draw(divout.center) to (w);
                \draw(one.center) to (z);
        \end{pgfonlayer}
        \begin{pgfonlayer}{eqlayer}
                \draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
                \draw[dashed, rounded corners](0.6, 0.3) rectangle (1.2, -1.3);
        \end{pgfonlayer}\begin{pgfonlayer}{background}
                        \draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
        \end{pgfonlayer}\end{tikzpicture}
}
\end{minipage}

\noindent
The right-hand side is a regular mono, though, and if the starting term graph is acyclic, so is also the arriving one.
Such an asymmetry between and left- and right-hand side falls in the current research about left-linear rules for 
adhesive categories, as pursued e.g. in~\cite{BaldanC0G24}.

>>>>

\noindent
Thus, $L$ and $R$ are the lift- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet they are not affected by the rule itself.
%
If a category is adhesive, and the arrows of the rules are monomorphisms, then the presence of a match ensures that the
two pushouts exist, hence a rewriting step can be performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{xxx} extend the core framework, still ensuring that if the arrow 
of the rules are in $\mathcal{M}$,
%and the match is in $\mathcal{N}$, 
then the classical results from the theory of graph 
transformation can still be lifted~\cite{xxx}. 
If only the left-hand side belongs to $\mathcal{M}$, the current theory is stil under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being adhesive can be a daunting task. For this reason,  sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their adhesivity has not yet been settled.

@@

\begin{restatable}{lemma}{kpp}\label{lemma:kern_pairs_pres_pullbacks}
	Let $f\colon X \to Y$ and $g\colon Z \to W$ be arrows admitting kernel pairs and suppose that the solid part of the four squares below is given. 
	%
	If the leftmost square is commutative, then there is a a unique arrow $k_h\colon K_f \to K_g$ making the other three commutative.
	\[\xymatrix@R=15pt{X \ar[r]^{h}\ar[d]_{f}& Z \ar[d]^{g} & K_f \ar@{.>}[r]^{k_h} \ar[d]_{\pi^1_f}& K_g \ar[d]^{\pi^1_g} & K_f  \ar@{.>}[r]^{k_h} \ar[d]_{\pi^2_f}& K_g \ar[d]^{\pi^2_g}& K_f  \ar@{.>}[r]^{k_h} \ar[d]_{(\pi^1_f,\pi^2_f)} & K_g \ar[d]^{(\pi^1_g, \pi^2_g)}\\ Y \ar[r]_t& W & X \ar[r]_{h} & Z & X\ar[r]_{h} &Z & X\times X \ar[r]_{h\times h}& X\times Z}\]

		Moreover, it holds
		\begin{enumerate}
			\item if $h$ is mono then $k_h$ is mono too;
			\item if the leftmost square is a pullback, then also the central too are so;
			\item if $h$ is mono and the first square is a pullback then the last one is a pullback too.
		\end{enumerate} 
\end{restatable}
%[Proof in \Cref{kpp-proof}]

@@

\noindent
\parbox{7.5cm}{\begin{restatable}{lemma}{mps}\label{prop:kerset}
	Suppose that in $\Set$ the cube in the diagram on the left is given, whose top and bottom faces are pushout, the left face a pulllback and $m\colon A\mto C$ is an injection. 
	Then it holds
	\begin{enumerate}
		\item the right face of the cube is a pullback too;
		\item the right square, made by the kernel pairs of the vertical arrows, is a pushout.
	\end{enumerate}
\end{restatable}}\hfill 
\parbox{6cm}{\xymatrix@C=10pt@R=5pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar@{>->}[dl]^{m} && B \ar@{>->}[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar@{>->}[rr]_{g} & & D }}

@@

Note that by hypothesis $(-)^\star$ preserves pullbacks, while $\pro$ is continuous by definition, thus the result below follows from  \Cref{thm:slice-functors} and \Cref{prop:com}.

\begin{corollary}\label[corollary]{prop:hypadh}
	 $\hyp$ is an adhesive category.
\end{corollary}

@@

\begin{definition}
	A \emph{hypergraph with equivalence} $\mathcal{G} = (E_\mathcal{G}, V_{\mathcal{G}}, Q_\mathcal{G}, s_\mathcal{G}, t_\mathcal{G}, q_\mathcal{G})$ is a 6-tuple such that $\mathcal{G} = (E_\mathcal{G}, V_{\mathcal{G}}, s_\mathcal{G}, t_\mathcal{G})$ is a hypergraph, $Q_\mathcal{G}$ is a set and $q_{\mathcal{G}}: V_{\mathcal{G}}\eto Q_{\mathcal{G}}$ is a surjection called \emph{quotient map}. 
	%
	A morphism $h\colon \mathcal{G\to H}$ is a triple $(h_E, h_V, h_C)$ such that the following diagrams commute
	\[\xymatrix@R=15pt{
		{E_\mathcal{G}}\ar[r]^{s_\mathcal{G}}\ar[d]_{h_E} & {V_{\mathcal{G}}^\star}\ar[d]^{h_V^\star} & {E_\mathcal{G}}\ar[r]^{t_\mathcal{G}}\ar[d]_{h_E} & {V_{\mathcal{G}^\star}}\ar[d]^{h_V^\star} & {V_\mathcal{G}}\ar@{>>}[r]^{q_\mathcal{G}}\ar[d]_{h_V} & {Q_{\mathcal{G}}} \ar[d]^{h_C} \\
		{E_\mathcal{H}}\ar[r]_{s_\mathcal{H}} & {V_{\mathcal{H}}^\star}	& {E_\mathcal{H}}\ar[r]_{t_\mathcal{H}} & {V_{\mathcal{H}^\star}}& {V_\mathcal{H}}\ar@{>>}[r]_{q_\mathcal{H}} & {Q_\mathcal{H}}
	}\]
	The category of hypergraphs with equivalence and their morphisms is denoted $\EqHyp$.

@@

\begin{definition}
	Let $\Sigma$ be an algebraic signature and $\mathcal{G}^{\Sigma}$ the hypergraph associated to it. A \emph{labelled hypergraph with equivalence} is a pair $(\mathcal{H}, l)$ where $\mathcal{H}$ is an object of $\EqHyp$ and $l$ a morphism $T(\mathcal{H})\to \mathcal{G}^\Sigma$ of $\hyp$. A \emph{morphism of labelled hypergraphs with equivalence} between $(\mathcal{H}, l)$ and $(\mathcal{H}', l')$ is an arrow $h\colon \mathcal{H}\to \mathcal{H}'$ such that $l= l'\circ T(\mathcal{H})$.

@@

In general, note that both a maximally and a minimally shared representation of an EGG exist, and for our 

@@

What about rewriting? The theory of $\mathcal{M}$-adhesivity ensures that if the rules are spans of 
arrows in $\mathcal{M}$, then we can lift all the standard properties
of the DPO approach.
% to the category at hand. 
However, the structure of rules that is advanced in~\cite{DetlefsNS05},
is given as a span $L \leftarrow L \rightarrow R$, where the first component is the identity, thus in $\pbc_\Sigma$,
while the second component may not belong to $\pbc_\Sigma$.

\noindent
\begin{minipage}[l]{.48\linewidth}
\vspace{.1cm}
Consider e.g. an EGG rule such $x / x \to 1$, from the introductory example in~\cite{WillseyNWFTP21}.
It can be modelled as the DPO rule aside, concisely represented as the arrow $L \rightarrow R$,
which can be applied with a match that is injective on equivalence classes to the legal EGGs
above.
\end{minipage}\hfill
\begin{minipage}[r]{.60\linewidth}
\xymatrix{        
 \begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
                \node[style=node](v1) at (-0.9, 0.3){};
                \node[style=node](v2) at (-0.9, -0.3){};
                \node[style=medium box] at (0, 0){$/$};
                \node[style=none](divfst) at (-0.3, 0.3){};
                \node[style=none](divsnd) at (-0.3, -0.3){};
                \node[style=none](divout) at (0.3, 0){};
                \node[style=node] (w) at (0.9, 0){};
        \end{pgfonlayer}        
        \begin{pgfonlayer}{edgelayer}
                \draw(v1) to (divfst.center);
                \draw(v2) to (divsnd.center);
                \draw(divout.center) to (w);
        \end{pgfonlayer}
        \begin{pgfonlayer}{eqlayer}
                \draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
                \end{pgfonlayer}\begin{pgfonlayer}{background}
                        \draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
                \end{pgfonlayer}
        \end{tikzpicture}
	\ar@{=>}[r] &
        \begin{tikzpicture}[baseline=(nil.center)]\begin{pgfonlayer}{nodelayer}
                \node[style=node](v1) at (-0.9, 0.3){};
                \node[style=node](v2) at (-0.9, -0.3){};
                \node[style=medium box] at (0, 0){$/$};
                \node[style=none](divfst) at (-0.3, 0.3){};
                \node[style=none](divsnd) at (-0.3, -0.3){};
                \node[style=none](divout) at (0.3, 0){};
                \node[style=node] (w) at (0.9, 0){};
                \node[style=small box] at (0, -1){$1$};
                \node[style=none](one) at (0.3, -1){};
                \node[style=node](z) at (0.9, -1){};
                \node[style=none](nil) at (0, -0.5){};
        \end{pgfonlayer}        
        \begin{pgfonlayer}{edgelayer}
                \draw(v1) to (divfst.center);
                \draw(v2) to (divsnd.center);
                \draw(divout.center) to (w);
                \draw(one.center) to (z);
        \end{pgfonlayer}
        \begin{pgfonlayer}{eqlayer}
                \draw[dashed, rounded corners](-1.2, 0.6) rectangle (-0.6, -0.6);
                \draw[dashed, rounded corners](0.6, 0.3) rectangle (1.2, -1.3);
        \end{pgfonlayer}\begin{pgfonlayer}{background}
                        \draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
        \end{pgfonlayer}\end{tikzpicture}
}
\end{minipage}

\vspace{.1cm}
\noindent
The right-hand side is a regular mono, though, and if the starting term graph is acyclic, so is also the arriving one.
Such an asymmetry between and left- and right-hand side falls in the current research about left-linear rules for 
adhesive categories, as pursued e.g. in~\cite{BaldanC0G24}.
